---
title: "GGplot practice"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

Loading in packages

```{r}
library(tidyverse)
library(here)
library(janitor)
```

## Ggplot basics

Read in lizard data
```{r}
lizards <- read_csv(here("data_tidy", "lizards.csv"))
```

### The essentials
- data
- asesthetics
- a geometry

```{r}
# ggplot essentials, 3 ways

# first (most basic)
ggplot(data = lizards,
       aes(x = total_length, y = weight)) + #essentials one and two
  geom_point() #essential three

#second; different syntax, same figure
ggplot(data = lizards) + 
  geom_point(aes(x = total_length, y = weight))

#third; nothing stays consistent, so could create figure with different types of data
ggplot() + 
  geom_point(data = lizards,
             aes(x = total_length, y = weight))

```

## Swap out geometries
```{r}
#bad example
ggplot(data = lizards,
       aes(x = total_length, y = weight)) +
  geom_line()
```

Some geometry types only need one axis
```{r}
ggplot(data = lizards, aes(x = total_length)) + #only need one variable!
  geom_histogram() 
```

Jiiterplots usually are best with one categorical and one continuous variable
```{r}
ggplot(data = lizards,
       aes(x = weight, y = common_name)) + #when you have a categorical varibale w/ different long names, see if you can plot it on the y axis instead (to make it easier to read)
  geom_jitter() #jitter spreads out in y direction
```

## Aesthetic mapping
### Updating based on a constant value
Don't need to use the 'aes()' function!

```{r}
ggplot(data = lizards, aes(x = weight)) + 
  geom_histogram(color = "orange", fill = "purple", alpha = 0.5, linetype = "dotted")
```

## Updating aesthetics based on variables
In that case, we need to use the aes() function

```{r}
ggplot(data = lizards,
       aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name),
             alpha = 0.3) 
#in order to use a variable from the data set, you need to be inside the aes() function; if you want to set aesthetics to a constant, you can do it outside of the function
```

# Themes
## Update axis labels
```{r}
ggplot(data = lizards, aes(x = weight, y = site)) + 
  geom_jitter(aes(color = common_name)) + 
  labs(x = "Weight (g)",
       y = "Study site",
       title = "Lizard weights across sies at Jornada",
       color = "Species") + #to change legend title, in labs() use the same argument that you're coloring by; in this case common_name is used to define color, so I change legend title by defining "color"
  theme_minimal()
```

## Reorder data
Default order is alphabetical, but you can also add function within fct_reorder() to tell it what to order your data/graph by

```{r}
lizard_counts <- lizards %>% 
  count(common_name) #creates a df with two columns: one for common_name and one for count!

ggplot(data = lizard_counts,
       aes(x = n, y = fct_reorder(common_name, n))) + #fct_reorder() puts it in order of highest count decreasing to lowest count; "factor is mapped to position"
  geom_col() + 
  labs(x = "Individuals (n)", 
       y = "Species common name")
```

## Advanced ggplot customization

```{r}
#if you want to do something with your axes, use the scale function!

ggplot(data = lizards,
       aes(x = total_length, y = weight)) +
  geom_point() +
  scale_x_continuous(breaks = c(0, 250, 500),
                     limits = c(0,500)) + #limit argument added a little bit more white space to ploot so you can actually see the 500 tick mark on the x axis 
  scale_y_continuous(breaks = seq(from = 0, to = 70, by = 10), limits = c(0, 70)) +
  labs(x= "Total Length (mm)",
       y = "Weight (grams)") +
  theme_minimal()

#to put on log scale, can use + scale_x_log10()

```


```{r}
#counting up the number of lizards caught per day
lizard_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  count(date)

#date-specific functions necessary because they're a different data format!
ggplot(data = lizard_counts,
       aes(x = date, y = n)) +
  geom_line() +
  scale_x_date(breaks = scales::breaks_width("2 years"), #changes the breaks of the dates
               labels = scales::label_date("%Y")) #%Y asks for the axis to be just the year (lowercase makes it just the last two digits)
```

### Customization of aesthetics with 'scale_' functions

```{r}
ggplot(data = lizards,
       aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) + 
  scale_color_gradient(low= "red", high="blue")

#color gradient with more than two colors
# + scale_color_gradient(colors = c(#vector with colors here))
  
#create a gradient of color bins
ggplot(data = lizards,
       aes(x = total_length, y = weight)) + 
  geom_point(aes(color = weight)) + scale_color_steps (low = "red", high = "blue",
                                                       breaks = seq(0, 80, 5))
```

```{r}
library(paletteer)
ggplot(data = lizards,
       aes(x = common_name, y = total_length)) + 
  geom_boxplot(aes(fill = common_name), 
               show.legend = FALSE) + 
  coord_flip() #coord_flip() can help you flip your axes around!
```

```{r}
#ordering species based on median length
lizard_mean <- lizards %>% 
  mutate(common_name = fct_reorder(common_name,
                                   total_length, .fun = median))

#using a palette that someone else created! via paletter
ggplot(data = lizard_mean,
       aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) + 
  scale_fill_paletteer_d(palette = "ggsci::default_gsea") + #last bit of this function dictates which kind of data you're using it for
  theme_minimal() + labs(x = "Total Length (mm)",
                         y = "Species") + 
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) #gets rid of the minor grid marks

#important to know that these are all layers, so if you put theme_minimal() after removing the grid lines, theme_minimal() will write over it and put them back in!
```

# Working with patchwork to creat compound figures
```{r}
library(patchwork)

#saving plots as objects
p1 <- ggplot(data = lizards,
       aes(x = total_length, y = weight)) + 
  geom_point(aes(color = weight)) + scale_color_steps (low = "red", high = "blue",
                                                       breaks = seq(0, 80, 5))

p2 <- ggplot(data = lizard_mean,
       aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) + 
  scale_fill_paletteer_d(palette = "ggsci::default_gsea") + #last bit of this function dictates which kind of data you're using it for
  theme_minimal() + labs(x = "Total Length (mm)",
                         y = "Species") + 
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) #gets rid of the minor grid marks
```

```{r}
(p1 | p2) & theme_minimal() #puts the two side by side!! and puts both of them in the minimal theme

# the | put plots next to each other, the / puts them on top of each other

(p1/p2) #worked!
(p1/p1/p1/p1)

#can do combinations of next to and on top of each other
p3 <- ((p1|p2)/p2)

ggsave("test.png", p3)

```

