---
title: "GGplot practice"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

Loading in packages

```{r}
library(tidyverse)
library(here)
library(janitor)
```

## Ggplot basics

Read in lizard data
```{r}
lizards <- read_csv(here("data_tidy", "lizards.csv"))
```

### The essentials
- data
- asesthetics
- a geometry

```{r}
# ggplot essentials, 3 ways

# first (most basic)
ggplot(data = lizards,
       aes(x = total_length, y = weight)) + #essentials one and two
  geom_point() #essential three

#second; different syntax, same figure
ggplot(data = lizards) + 
  geom_point(aes(x = total_length, y = weight))

#third; nothing stays consistent, so could create figure with different types of data
ggplot() + 
  geom_point(data = lizards,
             aes(x = total_length, y = weight))

```

## Swap out geometries
```{r}
#bad example
ggplot(data = lizards,
       aes(x = total_length, y = weight)) +
  geom_line()
```

Some geometry types only need one axis
```{r}
ggplot(data = lizards, aes(x = total_length)) + #only need one variable!
  geom_histogram() 
```

Jiiterplots usually are best with one categorical and one continuous variable
```{r}
ggplot(data = lizards,
       aes(x = weight, y = common_name)) + #when you have a categorical varibale w/ different long names, see if you can plot it on the y axis instead (to make it easier to read)
  geom_jitter() #jitter spreads out in y direction
```

## Aesthetic mapping
### Updating based on a constant value
Don't need to use the 'aes()' function!

```{r}
ggplot(data = lizards, aes(x = weight)) + 
  geom_histogram(color = "orange", fill = "purple", alpha = 0.5, linetype = "dotted")
```

## Updating aesthetics based on variables
In that case, we need to use the aes() function

```{r}
ggplot(data = lizards,
       aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name),
             alpha = 0.3) 
#in order to use a variable from the data set, you need to be inside the aes() function; if you want to set aesthetics to a constant, you can do it outside of the function
```

# Themes
## Update axis labels
```{r}
ggplot(data = lizards, aes(x = weight, y = site)) + 
  geom_jitter(aes(color = common_name)) + 
  labs(x = "Weight (g)",
       y = "Study site",
       title = "Lizard weights across sies at Jornada",
       color = "Species") + #to change legend title, in labs() use the same argument that you're coloring by; in this case common_name is used to define color, so I change legend title by defining "color"
  theme_minimal()
```

## Reorder data

```{r}
lizard_counts <- lizards %>% 
  count(common_name) #creates a df with two columns: one for common_name and one for count!

ggplot(data = lizard_counts,
       aes(x = n, y = fct_reorder(common_name, n))) + #fct_reorder() puts it in order of highest count decreasing to lowest count; "factor is mapped to position"
  geom_col() + 
  labs(x = "Individuals (n)", 
       y = "Species common name")
```

